Development of a Reinforcement Learning Enabled Cattle Tracker Prototype

### 2025-02-16 paper/main.tex
updated the main.tex and ref

There is an estimated 1.5 million beef cows in Alberta with a portion of those being free-range cattle who are allowed to graze on large pastures for extended periods of time.
Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
Reasearchers are interested feeding, rumination, and resting behaviour and past research has correlated it to sensor data from accelerometers \cite{unoldIoTBasedCowHealth2020}.
Monitors benefit from tracking devices that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
These trackers take the form of collars, leg bands, or ear tags and are equipped with sensors, batteries, and communication modules.
The installation of the trackers is a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
Given the size of the devices and the environment that they operate in, there is a limited power budget to balance contiunuous operation, transmission, and data collection.

The paradigm of Internet of Things (IoT) devices has emerged over the last two decades as a way to deploy sensors, communication modules, storage, compute, and intelligence throughout the environment.
The devices are small, low power, and have limited communication capabilities, but provide insight into the environment that was previously not possible.
These cattle trackers are IoT devices that replace labour intensive manual observation with automated systems that can monitor the herd 24-7 \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
IoT devices can operate as part of a network or as standalone units. 
For cattle tracking, a self-sufficient device that connects to the internet via cellular networks such as GSM, LTE, or even satellite services like Starlink is ideal, allowing farmers to monitor their herds remotely. 
However, these devices face power constraints due to the high data transmission rates required; solar panels can extend battery life, though their effectiveness depends on weather conditions and the cattle's orientation and location.

Reinforcement learning is a branch of machine learning that allows an agent to learn strategies to maximize a reward.
The agent learns by interacting with the environment and receiving rewards for those actions.
Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
The combination of reinforcement learning and IoT devices allows the device to make decisions without the need for human intervention.
The combination of AI and IoT is called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.

In this paper we present the development of a cattle tracker prototype capable of data collection for extended periods of time.
The device is designed to alternate between data collection, data transmission, and sleeping to conserve power.
It is not trivial to manually determine the optimal data collection strategy for the device under dynamic and complex conditions \cite{suttonReinforcementLearningIntroduction2020} so we propose the addition of reinforcement learning to automate the decision making process.
The main contribution of this paper is the development of a reinforcement learning agent that optimizes the power consumption of the device by allowing the agent to make decisions on when to transmit data, when to collect and store data, and when to do nothing.
The hardware prototype uses solar power to extend the power budget of the device and the agent is trained to take advantage of excess power availability to increase the data transmission rate.
The IoT device likely fits other use cases, but the focus in this paper is on the cattle tracking application due to the requirements of collecting uninterrupted data, occasionally needing to check in, needing to operate for long periods of time, and the potential for a mobile solar panel, which results in drastically different power generation.

@@

\begin{table}[h]
  \centering
  \begin{tabular}{l p{8cm} c}
      \toprule
      \textbf{Parameter} & \textbf{Description} & \textbf{Value} \\
      \midrule
      reward\_power\_loss & reward for losing power & 0.00  \\ 
      reward\_power\_multiplier & reward per power level at every time step & 0.0001  \\ 
      reward\_action\_0 & reward for doing nothing & -1.716 \\ 
      reward\_action\_1 & reward for collecting and storing data & -0.600 \\ 
      reward\_action\_2 & reward for collecting and transmitting data & -1.485 \\ 
      reward\_message\_count & reward for each message at transmission & 0.841 \\ 
      \bottomrule
  \end{tabular}
  \caption{Reward function parameters, their descriptions, and values extracted during hyperparameter tuning.}
  \label{tab:reward_parameters}
\end{table}

@@

\subsection{Hardware}
The hardware prototype is shown in \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
The detailed specifications are shown in table \ref{tab:hardware_inventory}.

>>>>

Alberta is home to an estimated 1.5 million beef cattle, many of which are free-range and graze on large pastures for extended periods of time.
Timely tracking of these animals is essential for farmers to assess herd health and monitor animal behaviors. 
For now, various Internet of Things (IoT)-based solutions have been developed to replace labor-intensive manual observation with automated, real-time monitoring \cite{unoldIoTBasedCowHealth2020}.
The devices are often small, low power, typically utilize collars, leg bands, or ear tags equipped with sensors and communication modules, and have limited communication capabilities, but provide insights into the environment that they are deployed in \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
However, their installation is a time consuming process, making long-term autonomous operation (months), without human intervention, a key requirement.
Additionaly, power limitations remain another significant challenge, as frequent data transmission depletes battery, necessitrating the design of energy-efficient solutions.
Solar panels can extend battery life, while their effectiveness relys heavily on weather conditions and the cattle's orientation and location.

Reinforcement learning is a branch of machine learning that allows an agent to learn strategies to maximize a reward.
The agent learns by interacting with the environment and receiving rewards for those actions.
The combination of reinforcement learning and IoT devices, also called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, offers a promising alternative by dynamically optimizing energy consumption through adaptive decision-making \cite{suttonReinforcementLearningIntroduction2020}. 
The TinyML paradigm, a subset of AIoT, focuses on machine learning on devices with highly constrained resources and is often limited to inference only\cite{rayReviewTinyMLStateoftheart2022}.
Overall, the addition of intelligence to the device allows it to make better decisions, however, on the other hand the addition of intelligence to the device increases the power consumption of the device.

In this paper, we present a reinforcement learning-enabled cattle tracker that autonomously manages power consumption by deciding when to transmit, store, or delay data collection. 
The device incorporates solar power to extend operational longevity, with the RL agent adapting to varying power availability. 
Our approach ensures uninterrupted monitoring while balancing energy efficiency and data quality. By leveraging RL, the proposed system outperforms static scheduling approaches, allowing for the collection of uninterrupted data, as well as occasionally check in, also making it a viable solution for long-term autonomous cattle tracking.

@@

\begin{table}[h]
  \centering
  \caption{Reward function parameters, their descriptions, and values extracted during hyperparameter tuning.}
  \begin{tabular}{l p{8cm} c}
      \toprule
      \textbf{Parameter} & \textbf{Description} & \textbf{Value} \\
      \midrule
      reward\_power\_loss & reward for losing power & 0.00  \\ 
      reward\_power\_multiplier & reward per power level at every time step & 0.0001  \\ 
      reward\_action\_0 & reward for doing nothing & -1.716 \\ 
      reward\_action\_1 & reward for collecting and storing data & -0.600 \\ 
      reward\_action\_2 & reward for collecting and transmitting data & -1.485 \\ 
      reward\_message\_count & reward for each message at transmission & 0.841 \\ 
      \bottomrule
  \end{tabular}
  \label{tab:reward_parameters}
\end{table}

@@

\subsection{Hardware}
The hardware prototype is shown in figure \ref{fig:prototype} and consists of an Arduino microcontroller, a GPS device, a solar array, a power distribution system, and a battery.
The detailed specifications are shown in table \ref{tab:hardware_inventory}.
